# Introduction to R

### A.  Let's Jump In
**RStudio - the console, the editor, the environment, the files, help**

- This is the R console. It's where you run commands in an interactive fashion. 
  - Error, warning, and informational output will also appear here.
- It's the same console you would get if you just run `R` on the command line.


- Working directory

  - Allows you to access/read/write files in a directory without having to give the whole directory path.  You can just give filenames relative to that directory.  Maybe right now this doesn't seem like a big deal.  But, as you do more work in R, you will see that it can get confusing to access files in lots of different places

  - Best practices keep all of your work for a single project in a single folder and make that your *working directory*  EVEN BETTER is to use an [RStudio project](https://support.rstudio.com/hc/en-us/articles/200526207-Using-Projects)

  - Ordinarily you would make a new project.  Today, we will check out this class project from git: 

    Upper right corner Project menu -> Version Control -> Git -> Repo URL: https://github.com/niaid/R_Intro -> Create Project

**Basic syntax in the console**

Variables are assigned using an arrow `<-` or an equals sign `=`.  `<-` is more common. Enter variable by itself to print, or use print statement.

```{r}
a <- 5
b <- 7
c_variable_1 <- a + b
print(c_variable_1)
```


Variable names have to follow some rules.
- allowed characters: alphanumeric, '.' and '_' - *no spaces*
- must start with a letter or '.'
- if it starts with '.', the second character cannot be a number
- of any length - easiest if you use meaningful names, so your code is readable by others

**Useful commands**

 - `ls()`, `rm`, `View`, `head`, `sessionInfo`

In an R session, objects are created in a workspace.

```{r}
ls()
rm(a,b)
ls()
```

- How do we get help?

1. Help pane with our mouse.
2. From the console:

```{r}
?help
```

- Help with errors
```{r}
foo(5,3)
```

- Search the internet!
   - stackoverflow.com - [Questions tagged with R](https://stackoverflow.com/questions/tagged/r)
   - [Cross Validated](https://stats.stackexchange.com/) - https://stats.stackexchange.com/
   - [Bioconductor Support Site](https://support.bioconductor.org/)
   - Package author's GitHub/GitLab/website may have an "Issues" section
   - [Common errors](http://varianceexplained.org/courses/errors/) 

**History**

By default R saves all the commands you run in the console in a file called *.Rhistory* in your working directory.  In RStudio, it shows up in the *History* pane.

**Saving your work**
By default the session is saved in a file called *.RData*.  This file is a special R binary format.  Because it starts with a '.', you won't see it if you do `ls` in your terminal, but it does show up in the RStudio Files browser.
RStudio will also ask you if you want to save your workspace when you quit (can turn this off in Tools > Global Options).

```{r, eval=F}
x <- 1:5
save.image()
rm(c_variable_1)
load('.RData')
```

Non-default save location and saving single objects (also in same R binary format)
```{r}
save.image("my.RData")
saveRDS(x, "x.rds")
y <- readRDS("x.rds")
```



**Writing a script**

- File -> New File -> R Script
- Paste commands from history
- Run from the command line.  The first line of your script (shebang):

```{bash}
#!/usr/bin/env Rscript
```
  - Then you can run in a terminal or command prompt:

  ```{bash}
  Rscript myscript.R
  ```
- Comments start with a `#` character.  These lines are not run.  Even better is to use `#'` - pound sign followed by apostrophe.
- Can use "knitr" button in RStudio to make a report.


### B. Basic Data Structures

#### Vectors
  
  - The basic data structures in R are vectors, data frames, matrices, lists, and functions.  Add-on packages may implement their own classes of data structures as well.
  - Vectors are the simplest data structures in R.
  - data types - character, numeric, factor, logical, `Inf`, `NA`, `NULL`
  - **numeric** vectors are collections of decimal numbers
  - **character** vectors are collections of strings which should be in single or double quotes to differentiate from variables in the workspace
  
```{r}
  num_vector <- c(7,3,4,6,2)
  char_vector <- c("volleyball", "  badminton", "netball", "tennis")
```
  
- **Accessing elements**

```{r}
num_vector[3]
num_vector[c(2,5)]
num_vector[2:5]
num_vector[-4]
num_vector[4] <- 37.5
```
  
- **Meta info about vectors**
```{r}
y_float <- c(5, sqrt(7), 3/21)
class(y_float)
typeof(y_float)
```

- **Head and tail**
```{r}
months <- month.name #constant in R
head(months)
tail(months)
```

#### Vector operations
Most operators/functions are vectorized - meaning they can take a vector of any length and perform the same operation element-wise.

*Example*: manipulate strings (useful for data cleanup) - substitute, trim
```{r}
gsub("a", "ae", char_vector)
char_vector <- trimws(char_vector)
```


*Example*: arithmetic operations
```{r}
x <- -1:12
x
5 + x/2
y <- c(0.2, 0.5)
x+y
```

- What happened with `x+y`??
  - shorter vectors are "recycled" in arithmetic operations, so *be careful* if this is not what you intend.
  - *Check* the length of vectors
```{r}
length(x)
length(y)
```


- **Lists**
  - A **list** is a general type of vector.  It can hold items from any data types or class, and the items can have different dimensions.
    - a data.frame is just a special kind of list where all the items are the same length
  - useful along with `lapply` to iterate over many items


- **Data frames (& Matrices)**
  
  - A *data frame* is a list of items which all have the same length.  In the most common case, these items are vectors and form columns of the data frame.
  - index elements
  - column and row names
  - data types of individual elements

### C.  File IO
**Reading in files**

Weber, Anna M., et al. “A Blue Light Receptor That Mediates RNA Binding and Translational Regulation.” *Nature Chemical Biology*, Aug. 2019, pp. 1–8. *www.nature.com*, doi:[10.1038/s41589-019-0346-y](https://doi.org/10.1038/s41589-019-0346-y). ([alt link](https://rdcu.be/bQhxo))

[Supplementary Dataset 1](https://static-content.springer.com/esm/art%3A10.1038%2Fs41589-019-0346-y/MediaObjects/41589_2019_346_MOESM3_ESM.xlsx) - 41589_2019_346_MOESM3_ESM.xlsx

[Supplementary Figure 6 - page 13](https://static-content.springer.com/esm/art%3A10.1038%2Fs41589-019-0346-y/MediaObjects/41589_2019_346_MOESM1_ESM.pdf#page=13)

- GUI: File -> Import Dataset -> 
- Paste command in script

**Writing tabular output**

- printing options

  - scipen
  
  > integer. A penalty to be applied when deciding to print numeric values in fixed or exponential notation. Positive values bias towards fixed and negative towards scientific notation: fixed notation will be preferred unless it is more than `scipen`digits wider.
  - in practice, if you don't want scientific notation, set:
  ```r
  options(scipen=999)
  ```
  - width - number of digits or characters before truncation/rounding
  ```r
  getOption("width")
  ```
  - Save data.frame to file

```{r}
?write.csv
write.csv(df, "mydata.csv", quote=FALSE)
?write_excel_csv ## readr package
```



### C.  Base and Add-on Packages
- base package
  - basic statistics - mean, median, sd, factorial
  - matrix operations
- how to install packages
- how to use functions from different packages
- install tidyr


### D. Plotting and Visualization

- [ggplot2](http://ggplot2.tidyverse.org) package
- ggplot2 (and the tidyverse) prefers long to wide data.
- "long" data format puts the data being analyzed/plotted in separate columns - one corresponding to each variable
- aesthetics

```{r}
vignette("ggplot2-specs")
```





### F.  Control Flow
- If statements - logical vectors, etc
- Functions
- Loops/Apply - simple for loops and also how to use lapply and apply
- debugging - `print`, `message`, `warn`, `stop`


