---
title: "Introduction to Rstudio (Part2)"
output:
  html_document:
    df_print: paged
  html_notebook: default
---

#### Use dplyr to manipulate data

##### 0. import data
```{r }  
library(readr)
BirdNest <- read_csv("BirdNest.csv")  # read data from csv file
# If you did not set work directory or would like to read in file from other folder instead of working directory, specify path above
 
```

##### 1. select 
To select columns or drop columns of a data frame, use select(). 

* Select desired variables
```{r}
#Select four columns: Length, Nesttype, Location, No.eggs from original data, return the first six rows.
library(dplyr)
select <- select(BirdNest, Length, Nesttype, Location, No.eggs)
head(select)
```

* Drop undesired variables  
```{r}
#Remove two columns: Species and Common from original data  
drop <- select(BirdNest,-(Species:Common))
head(drop)
```

<br />     
Note: You may want to set some criteria for variable inclusion when you have a lot of variables in your data.  

```{r echo=FALSE, result = 'asis'}
summary <- c("Select everything but ","Select range ",
      "Select columns whose name contains a character string",
      "Select columns whose name starts with a string",
      "Select columns whose name ends with a string",
      "Select columns whose name matches a regular expression",
      "Select columns whose names are in a group of names"
)
usage <- c("-",":","contains()","start_with()","ends_with()","matches()","one_of()")

table <- cbind(usage,summary)
library(knitr)
kable(table)

```

```{r}
# Select variables contain "nest" (default case insensitive)
head(select(BirdNest, contains("nest")))

```

<br />   
 
##### 2. filter  
Use **filter** to select rows that meet criteria. you may use %in% when you have specified levels that would like the variable to be filtered on, however, when your criteria is blurred, you may use "grepl", to roughly search pattern in the variable and return the rows meet criteria.    
```{r}
#Select rows with Length more than 30
filter(BirdNest, Length>30)

#Select rows with No.eggs more than 6 and Location in "decid"
filter(BirdNest, No.eggs>6 , Location %in% c("decid")) ## or Location=="decid" if just one level

```


<br />   
  
**Question:**
Could you filter the rows with levels contains "Jay" in Common column? (hint: use grepl function)

```{r}
filter(BirdNest ,grepl("Jay",Common))

```

<br />  

##### 3. mutate  
When you want to create new columns based on the values in existing columns, for example, do calculation using existing variables, we’ll use the dplyr function mutate().  

```{r}
head(mutate(BirdNest, ratio = Incubate/Totcare))
head(mutate(BirdNest, ratio = Incubate/Totcare, inverse = 1/ratio))
head(mutate(BirdNest, cumsum_total = cumsum(Totcare)))
head(mutate(BirdNest, nor_Nest = Nestling/mean(Nestling, na.rm=T))) # na.rm=T removes the missing value when calculating mean 
```

Note: create variables with criteria  

```{r echo=FALSE, result = 'asis'}
summary <- c("Elementwise minimum or maximum",
      "Cumulative minimum and maximum ",
      "Cumulative sum and product")
usage <- c("pmin(), pmax()","cummin(), cummax()","cumsum(), cumprod()")

table <- cbind(usage,summary)
library(knitr)
kable(table)

```
<br />  

##### 4. arrange  
Arrange the rows of your data based according to the preferred order in the specified variable.
```{r}
# order the data by ascending length
head(arrange(BirdNest, Length))
head(arrange(BirdNest, Length, No.eggs)) 
```

<br />  


##### 5. pipes  
Pipes can be used when you want to many things to the same data set. It takes the output of one function and send it directly to the next. Different layes can be added in the pipes. You will need to consider the order of adding the layers, because it needs to execute the analysis you plan and satisfy pipe logic.  

Note: ctrl+shift+m for the **%>%** symbol for Mac.  

```{r}
head(select(BirdNest, Length, Nesttype, Location, No.eggs))
# equals to
head(BirdNest %>% select(Length, Nesttype, Location, No.eggs))

head(filter(BirdNest, No.eggs>6 , Location %in% c("decid")))
# equals to
head(BirdNest %>% filter( No.eggs>6 , Location %in% c("decid")))

# add layer
BirdNest %>% 
  filter( No.eggs>6 , Location %in% c("decid")) %>% 
  select(Species, Common, No.eggs, Location)

```

<br />    

**Question:**
Could you output top 10 observations with largest length? 

```{r}
BirdNest %>%
  arrange(desc(Length) ) %>% 
  head(n = 10)

```
<br />  

##### 6. summarise  
When you want to create a summary across the data, summarise() function in dplyr package can be used. Generally, it often combines with group_by, which creates a summary by subgroups. 
```{r}
# to create a summery about average, variance of lengt, and count distinct egg color 
BirdNest %>% summarise(mean_length = mean(Length), var_Length = var(Length), n_distict_color = n_distinct(Color))  

# to create a summery respective to same fields above within each egg color.
BirdNest %>% 
  group_by(Color) %>% 
  summarise(mean_length = mean(Length), var_Length = var(Length), n_distict_color = n_distinct(Color)) 

```
<br />  

##### 7. group by  
group_by() and summarise together can create a split-apply-combine analysis. group_by() splits the data into groups, summarise() provides summary function in each group and the summary for each subgroups are combined and returned.   

Note:  
Adding multiple variables in group_by() will return a summary with grouping by adding order.
```{r}
# create summary about mean and variance of legnth, number of distinct location by color and nesttype
BirdNest %>% 
  group_by(Color,Nesttype) %>% 
  summarise(mean_length = mean(Length), num_obs = n(), n_distict_location = n_distinct(Location))
```
<br />  

**Take home question:  **

Could you create a summary about the largest ratio (ratio = Nestling / Totcare) by nest type (excluding "cavity" category) and present the result in a descending order?

```{r , echo=FALSE}
BirdNest %>% 
  filter(!grepl("cavity",Nesttype)) %>% 
  mutate(ratio = Nestling / Totcare) %>% 
  group_by(Nesttype) %>% 
  summarise(max_ratio = max(ratio, na.rm=T)) %>%
  arrange(desc(max_ratio)) 

```

